---
title: "Processing Panwar expression data"
author: "Zuhaib Ahmed, Gitta Ekaputeri, Anne-Christin Hauschild"
date: "12/06/2021"
update: "29/08/2022" 
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stringr)
library(data.table)
library(reshape2)
library(miRBaseConverter)
setwd("/data/miRNA/panwarData/")
```

Load the data
```{r}
tissues <- read.table("./miRNA_tissue_expression.txt", header = T, sep = "\t", check.names = F)
tissues[1:5,1:5]
```

Combine all the rows with the same mature miRNA ID
```{r}
tissues2 <- as.data.table(tissues[,-c(2:3)])
tissues2 <- tissues2[, lapply(.SD, sum, na.rm=TRUE), by = Mature.miRNA.ID, .SDcols=names(tissues2)[-1]]
tissues2 <- as.data.frame(tissues2)
row.names(tissues2) <- tissues2$Mature.miRNA.ID
tissues2 <- tissues2[,-1]
names(tissues2) <- str_replace(names(tissues2), "(^.*)-.*", "\\1") # Keep only the SRA accessions for the columns
tissues2[1:5,1:5]
```

### Update IDs
I'm going to use mirbase22 to update the miRNA IDs

```{r}
mrbse <- read.table("../mirbase/mature_homo-sapiens_dataframe.txt", header = T, sep = "\t")
mrbse <- mrbse[-c(1,2),] # Remove the first two rows that don't match up with mirbase and are the only duplicates, and are the only sequences with T's for some reason.
head(mrbse)
```

```{r}
newMirs <- miRNAVersionConvert(row.names(tissues2), targetVersion = "v22")
newMirs[690:700,]
```

Removal of NAs

```{r}
a<-which(is.na(newMirs$TargetName))
length(a)
unique(newMirs[a,"TargetName"])
unique(newMirs[a,"OriginalName"])
newMirs2 <- newMirs[-which(is.na(newMirs$TargetName)),]
dim(newMirs)
dim(newMirs2)
nrow(newMirs) - nrow(newMirs2)
```

Addition of sequences

```{r}
newMirs3 <- merge(mrbse, newMirs2, by.x = 1, by.y = 2, all.x = F, all.y = T)
newMirs3 <- newMirs3[,-4]
head(newMirs3)
```

Merge these mappings and sequence with the data

```{r}
tissues3 <- merge(newMirs3, tissues2, by.x = 3, by.y = 0, all.x = T, all.y = F)
tissues3 <- tissues3[,-1] # Remove the old mir ID names
tissues3[1:5,1:5]
```

### Annotate the samples with metadata

Use the file `mirmine_samples_to_keep.txt`. This file is derived from the `miRmine-info.txt` file from mirmine and outlines which samples we want to keep.

```{r}
meta <- read.table("./mirmine_samples_to_keep.txt", header = T, sep = "\t")
head(meta)
```

Create tissue+disease groups from this info

```{r}
meta$Group <- paste(tolower(str_replace_all(meta$Tissue_cp, "[^A-Za-z0-9]+", "\\.")), 
                    tolower(str_replace_all(meta$Disease_cp, "[^A-Za-z0-9]+", "\\.")), 
                    sep = "_")
meta
```

The `plasma_non.fatigued` comes from the same study in mirmine as the `plasma.fatigued` samples, which is different than the `plasma_normal` samples. In that study (`High-Throughput Sequencing of Plasma MicroRNA in Chronic Fatigue Syndrome/Myalgic Encephalomyelitis`), it's found that the `plasma.non.fatigued` samples were the normal controls. 
Replace `plasma_non.fatigued` with `plasma_normal`.

```{r}
meta$Group[which(meta$Group == "plasma_non.fatigued")] <- "plasma_normal"
```
Remove the `_normal` suffix

```{r}
meta$Group <- str_replace(meta$Group, "_normal", "")
```

### Quantile every column

For each column, assign each miRNA a value 1-5 depending on which of the 20th percentiles it falls into

```{r}
# Given a column, assigns a number to each element from 0-5. All
# 0s get a 0, and the rest get a value according to the 20th percentile
# that it falls in among the non-zero values.
quintize <- function(vec) {
  qntls <- c(0, quantile(vec[which(vec != 0)], 0.2*(1:4)))
  vec2 <- sapply(vec, function(x) {
    if (x == 0) return(0)
    else return(max(which(x > qntls)))
  })
  return(vec2)
}
```

```{r}
tissues4 <- apply(tissues3[,-c(1:2)], 2, quintize)
tissues4[1:5,1:5]
```

Now to combine the samples belonging to the same group. Combine them by taking the average. This corresponds with the binarization. If any of the samples have a non-zero value for a particular miRNA, then the average will be non-zero, indicating some expression. Conversely, if one particular tissue has a binary value of 1, then it means one of its samples had a non-zero value, which means that they'll have a non-zero average. So non-zero average if and only if the binary value is a 1. 

```{r}
uniqTissues <- unique(meta$Group)
tissues5 <- lapply(uniqTissues, function(x) {
  df <- as.data.frame(tissues4[,meta$Experiment.Accession[which(meta$Group == x)]])
  return(rowMeans(df))
})
names(tissues5) <- uniqTissues
tissues5 <- do.call(cbind, tissues5)
tissues5 <- cbind(tissues3[,1:2], tissues5)
tissues5[1:5,c(1:2,16:20)]
```

```{r}
numMat <- t(apply(tissues5[,-c(1:2)], 1, as.numeric))
hist(rowMeans(numMat), breaks = ncol(numMat)+1, main = "Average miRNA expression across all [averaged] groups")
hist(apply(numMat, 1, function(x) return(length(which(x != 0)))), breaks = ncol(numMat)+1, main = "Distribution of miRNA expression across all groups")
summary(rowSums(numMat))
```

### Ensure the terms are standardized

Ensure that all the group names appear in the ontology or the corrections file.

Get the ontology and correction files
```{r}
ont <- read.table("../ontology.txt", header = F, sep = "\t")
corr <- read.table("../corrections.txt", header = T, sep = "\t")
head(ont)
head(corr)
```

```{r}
trms <- unique(unlist(strsplit(names(tissues5)[3:ncol(tissues5)], "_"))) # Splits the composite terms that contain both tissue+disease
trms[-which(trms %in% union(corr[,1], unique(unlist(ont[,c(1,3)]))))] # Which terms aren't in the ontology or the corrected terms
```
This means that all terms have been accounted for (i.e. all terms in the datasets are either in the ontology or the corrections file (column 1)). 

Correct the terms that need to be corrected in the long data.

```{r}
trms <- unique(unlist(strsplit(names(tissues5)[3:ncol(tissues5)], "_")))
trms[-which(trms %in% unique(unlist(ont[,c(1,3)])))]
```
It seems `plasma`, `nasopharyngeal.carcinoma`, and `sperm` are the only term that is in the ontology file. It is in the corrections file though. So it's accounted for. 
Replace this with the standardized term.
Remove the `sperm` column from the data due to unclear synonyms like `semen`.

```{r}
if ("sperm" %in% names(tissues5)) {
  tissues5 <- tissues5[,-which(names(tissues5) == "sperm")]
}
```

Standardize terms

```{r}
names(tissues5)[3:ncol(tissues5)] <- sapply(names(tissues5)[3:ncol(tissues5)], function(z) {
  y <- strsplit(z, "_")[[1]]
  retVal <- sapply(y, function(x) {
    if (x %in% corr$currentTerm) {
      return(corr$correctedTerm[match(x, corr$currentTerm)])
    } else {
      return(x)
    }
  })
  return(paste(retVal, collapse = "_"))
})
```

Finalize the data


### Change data to long format

```{r}
data_long <- melt(tissues5)
names(data_long) <- c("miR", "Seq", "Tissue", "Value")
data_long$Canonical <- T # All sequences used here are in mirbase22 (I checked). So all sequences are canonical (i.e. T)
data_long$Source <- "Panwar"
data_long <- data_long[,c(1,2,5,6,3,4)]
data_long$Binary <- sapply(data_long$Value, function(x) if (x == 0) return(0) else return(1))
head(data_long)
```

Write data to files

```{r}
write.table(data_long, "panwar_longData.txt", row.names = F, col.names = T, sep = "\t", quote = F)
write.table(unique(data_long[,1:3]), "panwar_miRNAs.txt", row.names = F, col.names = T, sep = "\t", quote = F)
writeLines(as.character(unique(data_long$Tissue)), "panwar_tissues.txt")
```













