---
title: "Processing lorenzi expression data"
author: "Zuhaib Ahmed, Gitta Ekaputeri, Anne-Christin Hauschild"
date: "29/08/2022" 
output: html_notebook
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stringr)
library(miRBaseConverter)
library(reshape2)

root_path <- "../"
## Define Standard folders
generaldata_path <- paste(root_path, "/source", sep="")
data_path <- paste(root_path, "data/miRNA/lorenziData", sep="")
result_path <- paste(root_path, "data/miRNA_final", sep="")
mirbaseFile<- paste(root_path, "data/mirbase/mature_homo-sapiens_dataframe.txt", sep="")

setwd(root_path)
```

### Read counts data

Load and transpose counts file (Sample IDs as header and miRNA IDs as first column)
```{r}
oriCountsFile <- read.csv(file = paste(sep="", data_path,'/mature_counts_all.csv'))
countsFile <- as.data.frame(t(oriCountsFile)) # transpose oriCountsFile
countsFile <- janitor::row_to_names(countsFile, row_number=1)
countsFile <- tibble::rownames_to_column(countsFile, var="miR")
countsFile$miR <- chartr('.', '-', countsFile$miR) # replace '.' with '-' in miR
countsFile
```

### Update the IDs 

```{r}
update <- miRNAVersionConvert(countsFile$miR)
head(update)
length(unique(countsFile$miR))
length(unique(update[,"OriginalName"]))
length(unique(update[,"TargetName"]))

```

Fetch the mirbase22 IDs and sequences

```{r}
mrbse <- read.table(mirbaseFile, header = T, sep = "\t")
mrbse <- mrbse[-c(1,2),] # The first two rows don't match up with mirbase, are the only duplicates, and are the only sequences with T's for some reason. I'll remove them.
dim(mrbse)
head(mrbse)
```

Create microRNA Mapping table and Update Countfile miRNA Row Names
```{r}
mrbse2 <- merge(mrbse, update, by.x = 1, by.y = 2, all.x = F, all.y = T)
mrbse2 <- mrbse2[,-4]
dim(mrbse2)
head(mrbse2)
rownames(mrbse2)<- mrbse2[,"OriginalName"]
rownames(countsFile)<-countsFile[,1]
rownames(countsFile)<- mrbse2[rownames(countsFile), "Name"]
```


Create Row for each miRNA and Merge this with the data

```{r}
mergedData <- matrix(0, nrow = dim(mrbse)[1], ncol= 302)
mergedData <- data.frame(mergedData)
mergedData<- cbind(mrbse, mergedData)
colnames(mergedData)<- c("Name", "Seq", colnames(countsFile)[-1])
rownames(mergedData)<-mergeData[,1]
mergedData[rownames(countsFile), colnames(countsFile)[-1]]<- countsFile[,-1]
#mergedData <- merge(mrbse2, countsFile, by.x = 3, by.y = 1, all.x = T, all.y = F)
dim(mergedData)
head(mergedData)
#mergedData <- mergedData[,-1] # Remove the original (old) IDs
mergedData[1:5,1:5]
mergedData

```


### Quintize the data

For each column: assign each miRNA a value 1-5 depending on which of the 20th percentiles it falls into. If the value is 0, it remains a 0

```{r}
# Given a column, assign a number to each element from 0-5. All
# 0s get a 0, and the rest get a value according to the 20th percentile
# that it falls in among the non-zero values.
quintize <- function(vec) {
  qntls <- c(0, quantile(vec[which(vec != 0)], 0.2*(1:4)))
  vec2 <- sapply(vec, function(x) {
    if (x == 0) return(0)
    else return(max(which(x > qntls)))
  })
  return(vec2)
}
```

```{r}
mergedData[,3:ncol(mergedData)] = lapply(mergedData[,3:ncol(mergedData)], FUN = function(y){as.numeric(y)})
dim(mergedData)
quintizedData <- apply(mergedData[,-c(1:2)], 2, quintize)
quintizedData <- cbind(mergedData[,1:2], quintizedData)
quintizedData[1:5,1:5]
dim(quintizedData)
```

### Combine the replicates

Get the meta information

```{r}
meta <- read.table(paste(sep="", data_path,"/meta_celltype_tissue_sRNA.txt"), header = T, sep = "\t")
meta$Sample_ID <- gsub("BioSample: https://www.ncbi.nlm.nih.gov/biosample/", "", meta$X.Sample_relation) # extract only SAMN IDs from X.Sample_relation
meta
```


Create Disease, Tissue, and Group (Tissue_Disease) columns

```{r}
meta$Tissue <- str_replace_all(meta$X.Sample_source_name_ch1, " tissue", "")
meta$Tissue <- str_replace_all(meta$Tissue, "[^A-Za-z0-9]+", "\\.")
meta$Tissue <- str_replace_all(meta$Tissue, "\\.donor\\.[0-9]+", "")
meta$Tissue <- str_replace_all(meta$Tissue, "-", ".")
meta$Tissue <- str_replace_all(meta$Tissue, " ", ".")
meta$Tissue <- tolower(meta$Tissue)
meta$Tissue <- str_replace_all(meta$Tissue, "cb.cd34.lin.", "cb.cd34.lin")
meta$Disease <- "" #all normal
meta$Group <- apply(meta[,c(ncol(meta)-1,ncol(meta))], 1, function(x) paste(x[1], x[2], sep="_"))
meta$Group <- str_replace_all(meta$Group, "_", "")
head(meta)
```


Grouping: Combine replicates (samples that have the same type)

```{r}
uniqueTissues <- unique(meta$Group)
meta$Sample_ID %in% names(quintizedData)
meta <- meta[!(meta$Sample_ID == "SAMN13014236"),] # Sample does not exist in counts file

# Remove columns in quintizedData that does not exist in meta (that means the sample type is not tissue and not cell type)
names(quintizedData) %in% meta$Sample_ID
quintizedData2 <- quintizedData[,colnames(quintizedData) %in% meta$Sample_ID]

# Add Name and Seq columns
quintizedData2 <- cbind(quintizedData[,1:2], quintizedData2)


#quintizedData2 <- cbind(quintizedData$Seq,quintizedData2)
#quintizedData2 <- cbind(quintizedData$Name,quintizedData2)
#names(quintizedData2)[1] <- "Name"
#names(quintizedData2)[2] <- "Seq"
#names(quintizedData2) %in% meta$Sample_ID




combinedData <- lapply(uniqueTissues, function(x) {
  df <- as.data.frame(quintizedData2[,meta$Sample_ID[which(meta$Group == x)]])
  return(rowMeans(df))
})
names(combinedData) <- uniqueTissues
combinedData <- do.call(cbind, combinedData)
combinedData <- cbind(quintizedData2[,1:2], combinedData)
dim(combinedData)
```


### Ensure the terms are standardized
Ensure that all the group names appear in the ontology or the corrections file.

Get the ontology and correction files

```{r}
ont <- read.table(paste(sep="", generaldata_path,"/ontology.txt"), header = F, sep = "\t")
corr <- read.table(paste(sep="", generaldata_path,"/corrections.txt"), header = T, sep = "\t")
head(ont)
```

Aim to have 0 as a result, meaning all terms are either already in the ontology or the correction file. Or else: update correction file or ontology file, or both.

```{r}
trms <- unique(unlist(strsplit(names(combinedData)[3:ncol(combinedData)], "_"))) # Splits the composite terms that contain both tissue+disease
if (length(which(trms %in% union(corr[,1], unique(unlist(ont[,c(1,3)]))))) != 0){
  trms[-which(trms %in% union(corr[,1], unique(unlist(ont[,c(1,3)]))))] # Which terms aren't in the ontology or the corrected terms
} else {
  trms
}
```

Correct current terms that need to be corrected.
```{r}
colnames(combinedData) <- sapply(colnames(combinedData), function(z) {
  y <- strsplit(z, "_")[[1]]
  retVal <- sapply(y, function(x) {
    if (x %in% corr$currentTerm) {
      return(corr$correctedTerm[match(x, corr$currentTerm)])
    } else {
      return(x)
    }
  })
  return(paste(retVal, collapse = "_"))
})
dim(combinedData)
head(combinedData)
```

### Convert to long format
Add the Canonical column and Source column

```{r}
n<-dim(combinedData)[1]
data<- cbind(combinedData["Name"],combinedData["Seq"],Canonical=rep(T, n) , Source=rep("Lorenzi*", n) , combinedData[,3:dim(combinedData)[2] ])
colnames(data)<- c("Name", "Seq", "Canonical", "Source", colnames(combinedData)[-c(1,2)])
colnames(data)
dim(data)
head(data)

```


Convert to long format

```{r}
data_long <- melt(data, id.vars=c("Name", "Seq", "Canonical", "Source"))
```

Binarization
```{r}
data_long <- data_long[!is.na(data_long$value),]
data_long$Binary <- sapply(data_long$value, function(x) if (x == 0) return(0) else return(1))
names(data_long) <- c("miR", "Seq", "Canonical", "Source", "Tissue", "Scale", "Binary")
```

### Write to file

```{r}
write.table(data_long, paste(sep="", result_path,"/lorenzi_longData.txt"), sep = "\t", row.names = F, col.names = T, quote = F)
write.table(unique(data_long[,1:3]), paste(sep="", result_path,"/lorenzi_miRNAs.txt"), sep = "\t", row.names = F, col.names = T, quote = F)
writeLines(as.character(unique(data_long$Tissue)), paste(sep="", result_path,"/lorenzi_tissues.txt"))
```