---
title: "Processing MC Call expression data"
author: "Zuhaib Ahmed, Gitta Ekaputeri, Anne-Christin Hauschild"
date: "16/07/2021"
update: "29/08/2022" 
output: html_notebook
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(microRNAome)
library(stringr)
library(miRBaseConverter)
library(reshape2)
```

### Load the data

```{r}
setwd("./data/miRNA/mccallData/")



#data("microRNAome")
#mirnome <- assay(microRNAome)
#rm(microRNAome)
#mirnome[1:5,1:5]
```

We need to remove any rows that have a `/` in them. They refer to multiple IDs and there's no way to determine which to attribute the counts to. 

```{r}
mirnome2 <- mirnome[-grep("\\/", row.names(mirnome)),]
dim(mirnome)
dim(mirnome2)
```

We lost 16 rows. Not too bad.

### Fetch metadata

The supplementary files provided have the metadata for some of the samples, but not all of them. There's probably a way of getting the meta data from SRADb, but I don't know how to yet. So for now, I'll just use the supplementary files. In the future, I'll be able to get the meta data for the remaining samples.
The following chunk of code compiles all the sample data from the supplementary files. We're not including cell lines so I've excluded supplementary file 3.

```{r}
s5 <- read.table("./supp_gr.222067.117_Supplemental_Table_S5.csv", header = T, sep = "\t")
s5meta <- names(s5)[grep("[SED]RR", names(s5))]
s5meta2 <- data.frame(ID = str_replace(s5meta, "(.*)\\.([SED]RR[^\\.]+)\\.*.*$", "\\2"),
                      Meta = str_replace(s5meta, "(.*)\\.([SED]RR[^\\.]+)\\.*.*$", "\\1"))
#s5meta2$Meta <- tolower(str_replace_all(s5meta2$Meta, "[^A-Za-z0-9]+", "\\."))

s2 <- read.table("./supp_gr.222067.117_Supplemental_Table_S2.csv", header = T, sep = "\t")
s2 <- s2[,c(3,1)]
#s2$Cell.Type <- tolower(str_replace_all(s2$Cell.Type, "[^A-Za-z0-9]+", "\\."))
colnames(s2) <- c("ID", "Meta")
s4 <- read.table("./supp_gr.222067.117_Supplemental_Table_S4.csv", header = T, sep = "\t")
s4 <- s4[,c(2,1)]
#s4$Tissue <- tolower(str_replace_all(s4$Tissue, "[^A-Za-z0-9]+", "\\."))
colnames(s4) <- c("ID", "Meta")

meta <- do.call(rbind, list(s5meta2, 
                            s2[-which(s2$SRA.Number %in% s5meta2$SRANumber),],
                            s4[-which(s4$SRAnumber %in% s5meta2$SRANumber),]))
meta <- meta[order(meta$Meta),]
colnames(meta) <- c("ID", "Group")
head(meta)
```

### Update miRNA IDs to mirbase22

I need to ensure that the miRNAs are up to date as well as to get the canonical sequence from mirbase. 

```{r}
newIDs <- miRNAVersionConvert(row.names(mirnome2))
head(newIDs)
```

```{r}
newIDs[which(is.na(newIDs[,2])),]
```
There are 29 NA values. I'll remove them.
```{r}
newIDs2 <- newIDs[-which(is.na(newIDs[,2])),-3]
```

There are some duplicated mappings too

```{r}
newIDs2[union(which(duplicated(newIDs2[,2])), which(duplicated(newIDs2[,2], fromLast = T))),]
```
I'll remove these rows as well

```{r}
newIDs3 <- newIDs2[-union(which(duplicated(newIDs2[,2])), which(duplicated(newIDs2[,2], fromLast = T))),]
```

Replace the old miR IDs with the new ones.

```{r}
mirnome3 <- merge(newIDs3, mirnome2, by.x = 1, by.y = 0, all.x = T, all.y = F)
mirnome3 <- mirnome3[,-1]
mirnome3[1:5,1:5]
```

Get the sequences from mirbase22

```{r}
mrbse <- read.table("../mirbase/mature_homo-sapiens_dataframe.txt", header = T, sep = "\t")
mrbse <- mrbse[-c(1,2),] # Remove the first two rows that don't match up with mirbase and are the only duplicates, and are the only sequences with T's for some reason.
head(mrbse)
```

Merge this with the count data

```{r}
mirnome4 <- merge(mrbse, mirnome3, by.x = 1, by.y = 1, all.x = F, all.y = T)
mirnome4[1:5,1:5]
```

### Quantile every column

For each column, I'll assign each miRNA a value 1-5 depending on which of the 20th percentiles it falls into

```{r}
# Given a column, assigns a number to each element from 0-5. All
# 0s get a 0, and the rest get a value according to the 20th percentile
# that it falls in among the non-zero values.
quintize <- function(vec) {
  qntls <- c(0, quantile(vec[which(vec != 0)], 0.2*(1:4)))
  vec2 <- sapply(vec, function(x) {
    if (x == 0) return(0)
    else return(max(which(x > qntls)))
  })
  return(vec2)
}
```

```{r}
mirnome5 <- apply(mirnome4[,-c(1:2)], 2, quintize)
mirnome5 <- mirnome5[,which(colnames(mirnome5) %in% meta$ID)]

mirnome5 <- cbind(mirnome4[,1:2], mirnome5)
mirnome5[1:5,1:5]
```

Now to combine the datasets that belong to the same group

Firstly, the cnvention of this project is to lowercase all letters and replance any string of non-alphanumeric characters with a `.`.

```{r}
meta2 <- meta
meta2$Group <- tolower(str_replace_all(meta2$Group, "[^A-Za-z0-9]+", "\\."))
```

```{r}
uniqTissues <- unique(meta2$Group)
mirnome6 <- lapply(uniqTissues, function(x) {
  df <- as.data.frame(mirnome5[,meta2$ID[which(meta2$Group == x)]])
  return(rowMeans(df))
})
names(mirnome6) <- uniqTissues
mirnome6 <- do.call(cbind, mirnome6)
mirnome6 <- cbind(mirnome5[,1:2], mirnome6)
mirnome6[1:5,c(1:2,16:20)]
```

### Standardize the column names

The column names need to be in accordance with the ontology terms. Load the ontology and corrections files to ensure that all names are standardized

```{r}
ont <- read.table("../ontology.txt", header = F, sep = "\t")
corr <- read.table("../corrections.txt", header = T, sep = "\t")
head(ont)
head(corr)
```

```{r}
trms <- unique(unlist(strsplit(names(mirnome6)[-c(1,2)], "_"))) # Splits the composite terms that contain both tissue+disease
trms <- trms[-which(trms %in% union(corr[,1], unique(unlist(ont[,c(1,3)]))))] # Which terms aren't in the ontology or the corrected terms
trms
```

Remove synonyms.
```{r}
trms2 <- unique(unlist(strsplit(colnames(mirnome6)[-c(1:2)], "_")))
trms2 <- trms2[-which(trms2 %in% unique(unlist(ont[,c(1,3)])))]
setdiff(trms2, trms)
```



Correct tissue files.

```{r}
names(mirnome6)[3:ncol(mirnome6)] <- sapply(names(mirnome6)[3:ncol(mirnome6)], function(z) {
  y <- strsplit(z, "_")[[1]]
  retVal <- sapply(y, function(x) {
    if (x %in% corr$currentTerm) {
      return(corr$correctedTerm[match(x, corr$currentTerm)])
    } else {
      return(x)
    }
  })
  return(paste(retVal, collapse = "_"))
})
```

Now remove those columns that didn't show up in the ontology

```{r}
mirnome7 <- mirnome6[,-which(colnames(mirnome6) %in% trms)]
ncol(mirnome6) - ncol(mirnome7)
```
So 61 of the column groups couldn't be added. Perhaps they can be added later on.

### Convert data to long format

First add in a `Canonical` and `Source` column. All sequences here are canonical. So we can simply fill the column with `TRUE`.

```{r}
mirnome8 <- mirnome7
mirnome8$Canonical <- T
mirnome8$Source <- "McCall"
mirnome8 <- mirnome8[,c(1,2,ncol(mirnome8)-1,ncol(mirnome8),3:(ncol(mirnome8)-2))]
head(mirnome8)
```

```{r}
data_long <- melt(mirnome8)
data_long$Binary <- sapply(data_long$value, function(x) if (x == 0) return(0) else return(1))
names(data_long) <- c("miR", "Seq", "Canonical", "Source", "Tissue", "Value", "Binary")
head(data_long)
```

### Write to File

```{r}
write.table(data_long, "mccall_longData.txt", col.names = T, row.names = F, sep = "\t", quote = F)
write.table(unique(data_long[,1:3]), "mccall_miRNAs.txt", col.names = T, row.names = F, sep = "\t", quote = F)
writeLines(as.character(unique(data_long$Tissue)), "mccall_tissues.txt")
```

























