---
title: "cleanupSampleAnnotations"
author: "Zuhaib Ahmed"
date: "10/06/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stringr)
library(reshape2)
setwd("data/RNA/gtexData/")
```

Load annotations
```{r}
anotable <- read.table("sampleAnnotations.txt", header = T, sep = "\t", quote = "")
```
According to the data dictionary on the GTEx website, we need the `SAMPID` and `SMTSD` columns. These are the sample names and tissue of sample origin.
```{r}
anotable <- anotable[,c("SAMPID", "SMTSD")]
head(anotable)
```
The convention in this project is to take all tissues/diseases, change any string of non-alphanumeric chcaracters into `.` and make all letters lowercase. 
Change all `-` into `.` in the sample names because once I read in the data files, the header will contain `.`s instead of `-`s. 

```{r}
anotable$SMTSD <- tolower(str_replace_all(anotable$SMTSD, "[^A-Za-z0-9]+", "\\."))
anotable$SAMPID <- str_replace_all(anotable$SAMPID, "-", "\\.")
names(anotable) <- c("Sample", "Group")
head(anotable)
```
Split up the data file row-wise to handle each chunk separately, write each chunk to a separate file and then merge them all together

Read the first chunk which contains the header.

```{r}
chunk1 <- read.table("rawChunk_aa", header = T, sep ="\t")
```

```{r}
hdr <- colnames(chunk1)
head(hdr)
```

```{r}
uniqTissues <- unique(anotable$Group)
```
```{r}
processChunk <- function(chnk) {
  colnames(chnk) <- hdr
  chnk_b <- apply(chnk[,-c(1:2)], 2, function(s) {
    s <- sapply(s, function(z) if (z < 0.1) return(0) else return(1))
    return(s)
  })
  data <- lapply(uniqTissues, function(x) {
    vec <- anotable$Sample[which(anotable$Group == x)] # These two lines can be factored out since they're the same for every chunk.
    vec <- vec[which(vec %in% colnames(chnk_b))]
    df <- chnk_b[,vec]
    return(sapply(rowSums(as.data.frame(df)), function(y) {
      if (y < (0.2*ncol(df))) return (0)
      else return(1)
    }))
  })
  names(data) <- uniqTissues
  data <- do.call(cbind, data)
  data <- cbind(chnk[,1:2], data)
  return(data)
}
```
```{r}
chunk1_p <- processChunk(chunk1)
```

Read in the rest of the chunks
```{r}
chnkFiles <- list.files()
chnkFiles <- chnkFiles[grep("rawChunk_", chnkFiles)]
chnkFiles <- chnkFiles[-which(chnkFiles == "rawChunk_aa")] # 
chnkFiles
```

Now load and process the other chunks
```{r}
dataList <- lapply(chnkFiles, function(f) {
  print(f)
  chnk <- read.table(f, header = F, sep ="\t")
  chnk_p <- processChunk(chnk)
  return(chnk_p)
})
```

Consolidate all the data
```{r}
data_wide <- do.call(rbind, dataList)
data_wide <- rbind(chunk1_p, data_wide)
dim(data_wide)
```
```{r}
hist(rowSums(data_wide[,-c(1:2)]), breaks = ncol(data_wide)-1)
summary(rowSums(data_wide[,-c(1:2)]), breaks = ncol(data_wide)-1)
```
Filter by protein-coding genes.
Get up-to-date protein coding genes from the hgnc website.

```{r}
pcg <- read.table("../protein-coding_genes_hgnc.txt", header = T, sep = "\t", quote = "", fill = T)
length(intersect(pcg$symbol, data_wide$Description))
data_wide2 <- data_wide[which(data_wide$Description %in% pcg$symbol),]
dim(data_wide2)
```

Remove duplicated hgnc symbols.

```{r}
data_wide2 <- data_wide2[-union(which(duplicated(data_wide2$Description)), which(duplicated(data_wide2$Desc, fromLast = T))),]
dim(data_wide2)
```
Let's check the distribution now
```{r}
hist(rowSums(data_wide2[,-c(1:2)]), breaks = ncol(data_wide2)-1)
summary(rowSums(data_wide2[,-c(1:2)]), breaks = ncol(data_wide2)-1)
```
Convert this data into long format for saving

Remove the ensembl column
```{r}
data_long <- melt(data_wide2[,-1])
data_long$Source <- "GTEx"
data_long <- data_long[,c(1,4,2,3)]
names(data_long) <- c("Gene", "Source", "Tissue", "Value")
head(data_long)
```

```{r}
 write.table(data_long, "gtex_longData.txt", row.names = F, col.names = T, sep = "\t", quote = F)
 writeLines(as.character(unique(data_long$Gene)), "gtex_genes.txt")
 writeLines(as.character(unique(data_long$Tissue)), "gtex_tissues.txt")
```







