---
title: "Creating Tissue Mapping and Tissue Groupings"
author: "Zuhaib Ahmed / Anne-Christin Hauschild"
date: "16/06/2021"
update: "29/08/2022" 
output: html_document
---

```{r setup, include=FALSE}
#Linux
source_path <<- "./"
root_path <<- "../"


## Define Standard folders
data_path <- paste(root_path, "data/", sep="")
results_path <- paste(root_path, "results/", sep="")
## Define files:
ontologyfile<-paste(data_path, "ontology.txt", sep="")
allTissuesMirsFile<-paste(results_path, "allTissues_mirs.txt", sep="")
allTissuesGenesFile <- paste(results_path, "allTissues_genes.txt", sep="")
allTissuesNovelmirsFile <- paste(results_path, "allTissues_novelMirs.txt", sep="")
uncoveredTermsFile<-paste(results_path, "allUncoveredTerms.txt", sep="")

allTissueMappingsFile <- paste(results_path, "allTissueMappings_220904.txt", sep="")
allTissueGroupingsFile<- paste(results_path, "allTissueGroupings_220904.txt", sep="")

knitr::opts_chunk$set(echo = TRUE)
setwd(data_path)
```

### Functions

```{r}
## getAllAncestors Function
## description: Get Mapping for all disease and tissues to high level contexts
getAllAncestors <- function(n) {
  parents <- ont[grep(paste0("^", n, "$"), ont[,1]),3]
  if (length(parents) == 0) return(n)
  anc <- c(n)
  for (p in parents) {
    anc <- c(anc, getAllAncestors(p))
    #print(des)
  }
  anc <- unique(anc)
  return(anc)
}
getAllDescendants <- function(n) {
  childs <- ont[grep(paste0("^", n, "$"), ont[,3]),1]
  
  if (length(childs) == 0) return(n)
  des <- c(n)
  for (c in childs) {
    des <- c(des, getAllDescendants(c))
    #print(des)
  }
  des <- unique(des)
  return(des)
}
```


### Read in the ontology file.
```{r}
ont <- read.table(ontologyfile, sep = "\t", header = F, fill = T)
head(ont)
sum(ont[,4]=="BTO")

terms <- unique(unlist(ont[,c(1,3)])) # Get unique terms in the ontology
```

# Read in the `allTissues_*` files. This has all tissue+disease (i.e. contexts) used in this project. We want to determine a set of nodes in the ontology that can encompass all these terms
```{r}
allTissues_mirs <- readLines(allTissuesMirsFile)
allTissues_genes <- readLines(allTissuesGenesFile)
allTissues_novelmirs <- readLines(allTissuesNovelmirsFile)
allTissues <- unique(unlist(list(allTissues_novelmirs, allTissues_genes, allTissues_mirs)))
allTerms <- unique(unlist(strsplit(allTissues, "_")))
head(allTerms)
```

# Get ancestorList
```{r}
ancList <- lapply(allTerms, function(x) {
  return(getAllAncestors(x))
})
names(ancList) <- allTerms
head(ancList)
```

#Manually curated cutoff nodes.

```{r}
cutoffNodes <- c("connective.tissue", "gland", "viscus", "head", "bone.marrow",
                 "brain", "heart", "kidney", "bladder",
                 "lung", "bronchus", "immune.system", "blood", "serum", "lymph.node", "tonsil",
                 "male.reproductive.system", "female.reproductive.system",
                 "skeletal.muscle", "mediastinum", "retroperitoneum",
                 "trunk", "eye", "tongue", "bile.duct", "placenta", "limb",
                 "tuberculosis", "chronic.fatigue.syndrome", "marginal.zone.b.cell.lymphoma",
                 "tumor", "sarcoma", "malignant.mesothelioma", "carcinoma", "low.grade.glioma",
                 "cancer", "pheochromocytoma", "skin", "liver.cirrhosis", "bone.disease",
                 "cardiovascular.system", "endothelial.cell", "epithelial.cell", "nervous.system",
                 "embryonic.structure", "trachea", "parkinsons.disease")
```

# Determine Term Coverage

```{r}
termCoverage <- function(n) {
  names(which(sapply(getAllDescendants(n), function(x) if (x %in% allTerms) return(T) else return(F))))
}

head(allTerms)
uncoveredTerms <- setdiff(allTerms, unique(unlist(lapply(cutoffNodes, termCoverage))))
head(uncoveredTerms)
write.table(uncoveredTerms, uncoveredTermsFile, row.names = F, col.names = F, sep = "\t", quote = F)
```

# Determine Missing Terms

```{r}
missingFromMirs <- intersect(uncoveredTerms, unique(unlist(strsplit(allTissues_mirs, "_"))))
missingFromNovel <- intersect(uncoveredTerms, unique(unlist(strsplit(allTissues_novelmirs, "_"))))
missingFromGenes <- intersect(uncoveredTerms, unique(unlist(strsplit(allTissues_genes, "_"))))
missingFromMirs
missingFromNovel
missingFromGenes
```

#Create the data frame that maps the low-level to high-level terms
```{r}
head(allTissues)
lowToHighDF <- lapply(allTissues, function(x) {
  splt <- strsplit(x, "_")[[1]]
  if (length(splt) == 2) {
    vec <- cutoffNodes[which(cutoffNodes %in% getAllAncestors(splt[2]))]
  } else {
    vec <- cutoffNodes[which(cutoffNodes %in% getAllAncestors(splt))]
  }
  return(data.frame(lowLevel = rep(x, length(vec)), highLevel = vec))
})
lowToHighDF <- do.call(rbind, lowToHighDF)
head(lowToHighDF)

```

#Creating the tissue grouping file that indicates whether the term is normal or diseased
```{r}
highLevelTerms <- unique(lowToHighDF$highLevel)
tissueGroupings <- data.frame(Tissue = highLevelTerms, Group = sapply(highLevelTerms, function(x) {
  anc <- getAllAncestors(x)
  if ("whole.body" %in% anc) return("normal")
  else return("disease")
}))
row.names(tissueGroupings) <- 1:nrow(tissueGroupings)
head(tissueGroupings)
```

#Writing Tissue mapping and Tissue grouping files.

```{r}
write.table(lowToHighDF, allTissueMappingsFile, row.names = F, col.names = T, sep = "\t", quote = F)
write.table(tissueGroupings, allTissueGroupingsFile, row.names = F, col.names = T, sep = "\t", quote = F)
```